Optimal Decomposition-Recombination Plans for 2D Well-Constrained Graphs

### 2014-12-03 doc/GMP2015/s03_drp.tex
Changes to paper.

% --------------------------------------------------------------
%                       Initialize
% --------------------------------------------------------------

\documentclass[11pt]{article}

% \usepackage{multicol}
\usepackage{ulem}
\usepackage{comment}
\newcommand{\todo}[1]{\textsc{{\color{red}#1}}}

\usepackage{personalNotes}
% contains special macros for math environment:
%     unit, p, set, abs, floor, ceil, vec, norm, inner, kernel
% contains useful all-purpose macros:
%     n, lessSeparation, append

\usepackage{tikz}
\usetikzlibrary{shapes,snakes}
\newcommand{\tedge}[5]{\draw[#3] (#1)-- node[e, #5] (e#4) {#4} (#2)}

% --------------------------------------------------------------
%                    Beginning of Document
% --------------------------------------------------------------

\begin{document}
% \begin{multicols}{2}

% --------------------------------------------------------------
%                           Title
% --------------------------------------------------------------

\pntitle {Decomposition Recombination Plans for 2D Well-Constrained Graphs}
\pnauthor{Troy Baker}
\pndate  {May 13, 2014}

\pnmaketitle
\pnmakefooter

% --------------------------------------------------------------
%                           Body
% --------------------------------------------------------------

\section{Introduction}
\todo{
Other sections...
\begin{itemize}
    \item Intro and motivation (previous work, where this work leads, examples)
    \item Preliminaries
    \item DR-Plan (Canonical) + algo
    \item Recombination + algo (refer to other paper and make 2D)
    \item Solution space navigation
    \item Conclusion
\end{itemize}
}

\section{Preliminaries}

\subsection{Definitions}
% ISSUE: My definitions don't prevent non-positive densities... This means things aren't guaranteed to be connected...

\begin{definition}[Graph]
Graph $G=(V,E,w_V,w_E)$ is a set of vertices, $V$, and a set of edges, $E$, defined as a tuple of vertices from $V$. Additionally, there are two weight functions, $w_V: V \to \mathbb{R}^+$ and $w_E: E \to \mathbb{R}^+$.
\end{definition}

\begin{definition}[Density]
The density of a graph is defined as:
\[d(G) = \sum_{e\in E}{w_E(e)} - \sum_{v\in V}{w_V(v)}\]
% \[d(G) = \max\set{0,\sum_{v\in V}{w_V(v)} - \sum_{e\in E}{w_E(e)}}\]
\end{definition}

\begin{definition}[Independent or Sparse Graph]
\todo{Needs definition.}
\end{definition}

\begin{definition}[Over-constrained Graph]
Graph $G=(V,E,w_V,w_E)$ is over-constrained if, given constant $k$, there exists some induced subgraph $S\subseteq G$ such that $d(S) > k$. (where S is not trivial... This seems weird because you could have an over-constrained under-constrained graph...)
\end{definition}

\begin{definition}[Well-constrained Graph]
Graph $G=(V,E,w_V,w_E)$ is well-constrained if, given constant $k$, $d(G) = k$ and for all induced subgraphs $S\subseteq G$, (1) $d(S)\leq k$, or (2) given a set of trivial graphs $T$, $S$ is isomorphic to one of the graphs in $T$ (i.e.\ $S$ is trivial).
\end{definition}

% \begin{definition}[Well-constrained Graph]
% Graph $G=(V,E,w_V,w_E)$ is well-constrained if, given constant $k$ and a set of trivial graphs $T$, the graph satisfies (1) $d(G)=k$ and (2) all subgraphs are either under-constrained, well-constrained, or trivial (i.e.\ are isomorphic to one of the graphs in $T$).
% \end{definition}

\begin{definition}[Rigid Graph]
Graph $G=(V,E,w_V,w_E)$ is rigid if, given constant $k$, there exists some spanning subgraph $S\subseteq G$ such that $S$ is well-constrained.
\end{definition}

\begin{definition}[Under-constrained Graph]
Graph $G=(V,E,w_V,w_E)$ is under-constrained if, given constant $k$, $G$ is not rigid.
\end{definition}

% \begin{definition}[Under-constrained Graph]
% Graph $G=(V,E,w_V,w_E)$ is under-constrained if, given constant $k$, there exists some subgraph $G'\subseteq G$ such that $d(G') < k$.
% \end{definition}

\begin{definition}[Trivial Graph]
A trivial graph is ill-defined in the general case. The only strict requirements are: (1) it must be an over-constrained graph and (2) all of its subgraphs are also trivial.

\todo{Maybe leave this out until it's needed later?}
In the familiar geometric cases of $d$-dimensional space, all vertex weights are $d$, all edge weights are $1$, and constant $k= -{{d+1}\choose{2}}$. Trivial graphs for 2D would be a single vertex. Trivial graphs in 3D would be a vertex and an edge (2 vertices with an edge between). Etc. These trivial graphs capture the notion of the rotational symmetry that exists in geometric spaces.
\end{definition}

\begin{definition}[Decomposition Recombination (DR-) Plan]
The DR-plan of graph $G$, $DRP(G)$, is defined as a tree that has the following properties:
\begin{enumerate}
    \item The root of the tree `contains' $G$.
    \item For all nodes, $C$, that `contain' nontrivial graphs, its $N$ children, $C_1, \ldots, C_N$, are trivial or well-constrained vertex-maximal proper subgraphs of that node $C$.
    \item The vertex set of $\bigcup_{i=1}^N{C_i}$ is the vertex set of $C$.
    \item A node with a trivial subgraph is a leaf.
\end{enumerate}
% It can also be described recursively as: the root is $G$, its children are the trivial subgraphs and the DR-plans of its well-constrained vertex-maximal proper subgraphs whose union is $G$ itself.

An equivalent DAG can be constructed from this tree such that all nodes containing the same subgraphs are combined into one, with all edges preserved.

Note that nodes will be referred to interchangeably as ``the node that contains the (sub)graph $C$'' and as simply ``$C$''.
\end{definition}

\begin{definition}[Complete DR-Plan]
The complete DR-plan of graph $G$, $CompleteDRP(G)$, is the unique DR-plan but with a modified rule number 2: the children are \textit{all} of the trivial or well-constrained vertex-maximal proper subgraphs of that node $C$. By this, rule 3 is implicit for a complete DR-plan.
\end{definition}

\begin{definition}[Optimal DR-Plan]
The optimal DR-plan of graph $G$, $OptimalDRP(G)$, is the DR-plan that minimizes the maximum fan-in over all nodes in the tree. It is not necessarily unique.
\end{definition}

\subsection{Notation}
$G$ is taken to be a graph with values $(V,E,w_V,w_E)$. To keep problems meaningful we assume that $G$ is connected, as the methods discussed are used to solve systems of equations to establish linkages. It is not useful to solve for two bodies simultaneously if there are no constraints between them. Furthermore, to limit the scope of the paper, $G$ will always be taken to be well-constrained.
% THIS MEANS NO EDGES MAY BE LEFT OUT, ALSO.

$Idc(G,X)$ is the graph induced on $G$ with the vertex set $X\subseteq V$. This can also be overloaded such that, using graph $H=(W,F)$, $Idc(G,H)=Idc(G,W)$.

$C$ is an arbitrary graph in a node in $CompleteDRP(G)$. $C_i$ is the $i^{\text{th}}$ child of $C$. By definition of a DR-plan, it is implied that $C_i$ is a well-constrained vertex-maximal proper subgraph.

\section{The Canonical Optimal DR-Plan}
\begin{figure}
\begin{center}
\begin{tikzpicture}
    \tikzstyle{v}=[draw, circle, minimum size=0.75cm]
    \tikzstyle{e}=[]

@@

%     \tedge{v1}{v2}{solid}{}{};
%     \tedge{v1}{v5}{solid}{}{};
%     \tedge{v2}{v3}{solid}{}{};
%     \tedge{v2}{v6}{solid}{}{};
%     \tedge{v3}{v4}{solid}{}{};
%     \tedge{v3}{v7}{solid}{}{};
%     \tedge{v4}{v8}{solid}{}{};
% \end{tikzpicture}
\end{center}

\caption{A well-constrained graph whose nodes are subgraphs labeled with their density. Below are its 3 optimal DR-Plans (partially complete). The center vertex has weight $k$ and is well-constrained itself. The radial nodes have weights $x,y,z \neq k$ with equivalent edge weights, making any union between $k$ and the others well-constrained. \todo{Make this figure more clear.}}
\label{3-plans}
\end{figure}

@@

\begin{figure}
\begin{center}
\begin{tikzpicture}
    \draw[rotate=0]  (0,0) ellipse (1.5cm and .3cm);
    \draw[rotate around={-59:(.7,1.2)}] (.8,1.1) ellipse (1.5cm and .3cm);
    \draw[rotate around={59:(-.7,1.2)}] (-.8,1.1) ellipse (1.5cm and .3cm);

@@

    % \tedge{l}{bl}{solid}{}{};
    % \tedge{bl}{b}{solid}{}{};
    % \tedge{r}{br}{solid}{}{};
    % \tedge{br}{b}{solid}{}{};
\end{tikzpicture}
\end{center}

@@

\section{Future Work}
While this paper fleshes out the case of a well-constrained input graph, little work has been done on the over-constrained graph. This has to do with the fact that finding a DR-Plan is NP-complete. But is there some algorithm that will find an approximation within a reasonable factor?

Under-constrained is not worth studying in itself, because it will either be that all subgraphs are under-constrained or trivial or that some subgraphs are well-constrained or over-constrained. In this way, it is easily reduced to the other problems.
\\

The current method for finding $DRP(G)$ is the Frontier algorithm. Is there a way to adapt the Pebble Game algorithm? It is used to detect rigidity, but there is no obvious way to use it to find wcvmps (other than cutting out one node at a time).

Or maybe, instead of these top-down algorithms, there is some bottom-up algorithm?

Useful for glass networks.

\begin{comment}
\section{Formal Conjectures}

\subsection{Conjecture 1}

\begin{conjecture}
% Take $G$ to be well-constrained. Then, for all nodes $C$ in $DRP(G)$, either $\forall i,j (C_i \cap C_j \leq 1)$ or $\forall i,j (C_i \cap C_j \geq 2) \wedge (C_i \cup C_j = C)$.

% or...

Take $G$ to be well-constrained. The operator $\cap_V$ returns the count of the vertices in the intersection of two graphs. Then, for all nodes $C$ in $DRP(G)$, exactly one of the following is true:
\begin{align*}
    \forall i,j & (C_i \cap_V C_j = 1) \wedge \p{\bigcup_{k\in\set{1,\ldots,N}\setminus\set{i}}{C_k} \neq C} \text{ or} \\
    \forall i,j & (C_i \cap_V C_j \geq 2) \wedge (C_i \cup C_j = C)
\end{align*}

That is, either any two children share a single vertex and all of them are needed to form the parent node, or there is some significant overlap of 2 or more vertices and because of that only 2 children are needed at node $C$ to satisfy the definition of the DR-plan.

This implies that in the first scenario all proper vertex-maximal rigid subgraphs of $C$ must be in tree $DRP(C)$, and that in the second scenario only 2 such subgraphs need to be in the tree.

$C_i \cap_V C_j$ cannot be $<1$ because 0 is disjoint but $G$ is taken to be well-constrained (negative and non-integral values are meaningless).
\end{conjecture}

\begin{conjecture}
$DRP(G)$ is ``unique'' modulo choice of children. Which means that the choice of children cannot impact the solution of the framework.
\end{conjecture}

\end{comment}

\begin{comment}
\pnplaceholder
Notes:
Edges do not appear as nodes of the DR plan, although vertices do (as leaves). In general, the children of a node are well-constrained proper maximal subgraphs of the parent, whose vertex set union gives the parent's vertex set.

We are trying to formalize here the concept of a canonical DR-plan of a well-constained graph; i.e, whether the non-uniqueness of the decomposition - in the case that some pair of proper vertex maximal rigid subgraphs have a nontrivial intersection - makes any real difference.

Conjecture 1 (should be easy to prove):
Let $C = (V,E)$ be well-constrained and $C_1 =(V_1,E_1),C_2 = (V_2,E_2)$ any two proper vertex-maximal well-constrained subgraphs with $V = V_1\cup V_2$ and $|V_1\cap V_2| \ge 2$ (called a nontrivial intersection). Then the subgraph $S_{1,2}$ induced by $V_1\cap V_2$ must be well-constrained.

Here are corollary facts that should be easy to prove:
(1) There are no edges between $V_1 \setminus (V_1\cap V_2)$ and $V_2 \setminus (V_1\cap V_2)$.
(2) There is no rigid subgraph induced by a proper, nonempty subset of vertices of   $V_i \setminus (V_1\cap V_2)$ and a subset of $V_1\cap V_2$ of size at least 2.

Corollary of Conjecture 1 and resulting facts:
$C_1\cap C_2$ is a proper vertex maximal rigid subgraph of $C_i$. Either there is a unique second proper vertex maximal rigid subgraph of $C_i$ that intersects  $C_1\cap C_2$  nontrivially (on an edge), or every pair of proper vertex maximal rigid subgraphs of $C_i$ intersect trivially.

This would imply the DR-plans of $C_1$ and $C_2$ will be unique and have a common child in the DR-plan of $C_1\cap C_2$.

Conjecture 2: Let $C = (V,E)$ be well-constrained and $C_1 =(V_1,E_1),C_2 = (V_2,E_2)$ any two proper vertex-maximal well-constrained subgraphs with $V = V_1\cup V_2$ and $|V_1\cap V_1| \ge 2$. $C_3 =(V_3,E_3),C_4 = (V_4,E_4)$ any two proper vertex-maximal well-constrained subgraphs with $V = V_3\cup V_4$ and $|V_3\cap V_4| \ge 2$. Then both DR-plans where $C_1$ and $C_2$ are taken as children or where $C_4$ and $C_3$ are taken as children will have the same max fan-in. In particular, either DR-plan will contain $\bigcap\limits_i C_i$ as a node - at the 2nd level below $C$.
\end{comment}

% Lemma 1: if the graph is well-constrained and any two proper vertex maximal subgraphs have an overlap of only one vertex, then all of the subgraphs will have 1 or 0.

% Overlap of one vertex between any $C_i$ and $C_j$ means that $C\neq C_i\cup C_j$ because we know $C$ is well-constrained and the graph $C_i\cup C_j$ would be able to pivot around the single vertex.

% Need to define trivial as a graph that, were it only accounting for vertices, would be over-constrained. I.e.\ a single vertex in 2D. Or a vertex and an edge in 3D.

% \begin{theorem}
% phrasing...
% \end{theorem}

% \begin{proof}
% By definition, any subgraph of a well-constrained graph will either be (1) under-constrained, (2) well-constrained itself, or (3) an over-constrained graph, referred to as trivial. We will analyze all 3 possibilities in turn, in the case of a union of any two nontrivial well-constrained vertex-maximal proper subgraphs.

% (1) If the union results in an under-constrained subgraph, then the intersection of the two original subgraphs must be an over-constrained trivial subgraph (can be shown with simple arithmetic). As such, there must be additional vertex-maximal proper subgraphs, all of whose intersections with one another are over-constrained trivial subgraphs (including $\emptyset$). All of these edge-disjoint subgraphs form a complex structure that is well-constrained and must be the graph itself (otherwise we did not have vertex-maximal subgraphs to begin with).

% (2) If the union results in a well-constrained subgraph, it must be the entire graph. If this were not the case, the two original subgraphs could not have been vertex-maximal proper because we would have found a larger well-constrained subgraph containing them; namely, their union or something containing it would have been larger.

% More can be said about the intersection of these two subgraphs, $C_1=(V_1, E_1)$ and $C_2=(V_2, E_2)$. It cannot be over-constrained, otherwise the union could not be well-constrained (simple arithmetic, see the first case). The intersection also cannot be under-constrained. If $C_1 \cap C_2$ were under-constrained, then for $C_1$ to be well-constrained you would need to add more constraints (edge weight) than degrees of freedom (vertex weight) to the intersection. The same is true for $C_2$. If both of these were satisfied, then the union of $C_1$ and $C_2$ would result in an over-constrained nontrivial graph. This is a contradiction, because we have shown the union to be the graph itself, which was defined to be well-constrained. Therefore, the intersection must be well-constrained.

% (3) It is impossible for the union of 2 nontrivial subgraphs to be trivial.
% \end{proof}

% \begin{corollary}
% There is no rigid subgraph induced by a proper, nonempty subset of vertices of $V_i \setminus (V_1\cap V_2)$ and a subset of $V_1\cap V_2$ of size at least 2.
% \end{corollary}

% \begin{proof}
% Since $C_1\cup C_2$ is the entire graph,

% well... no. if $V_i$ contains $V_1 \setminus (V_1\cap V_2)$ or $V_2 \setminus (V_1\cap V_2)$, then you will induce a well-constrained graph.
% \end{proof}

% \begin{corollary}
% $C_1 \cap C_2$ is a well-constrained vertex-maximal proper subgraph of every $C_i$.
% \end{corollary}

% \begin{proof}
% First, it must vertex-maximal proper subgraph of $C_1$ and $C_2$. Say that there was some larger proper subgraph in $C_1$ that was well constrained. If this were true, then $C_2$ was not vertex-maximal because it could have included that entire subgraph and remained well-constrained. The same is true vice versa.

% need to show that that intersection is going to end up being a vertex-maximal subgraph of both of the original subgraphs. As a result, in the second case it doesn't matter which 2 you took.

% --------------------------------------------------------------
%                        End of Body
% --------------------------------------------------------------

% \end{multicols}
\end{document}

% --------------------------------------------------------------
%                       End of Document
% --------------------------------------------------------------

>>>>

\section{Decomposition-Recombination Plan}

\begin{figure*}\centering
\begin{tikzpicture}
    \tikzstyle{v}=[draw, circle, minimum size=0.75cm]
    \tikzstyle{e}=[]

@@

%     \tedge{v1}{v2}{solid}{}{};
%     \tedge{v1}{v5}{solid}{}{};
%     \tedge{v2}{v3}{solid}{}{};
%     \tedge{v2}{v6}{solid}{}{};
%     \tedge{v3}{v4}{solid}{}{};
%     \tedge{v3}{v7}{solid}{}{};
%     \tedge{v4}{v8}{solid}{}{};
% \end{tikzpicture}

\caption{A well-constrained graph whose nodes are subgraphs labeled with their density. Below are its 3 optimal DR-Plans (partially complete). The center vertex has weight $k$ and is well-constrained itself. The radial nodes have weights $x,y,z \neq k$ with equivalent edge weights, making any union between $k$ and the others well-constrained. \todo{Make this figure more clear.}}
\label{3-plans}
\end{figure*}

@@

\begin{figure}\centering
\begin{tikzpicture}
    \draw[rotate=0]  (0,0) ellipse (1.5cm and .3cm);
    \draw[rotate around={-59:(.7,1.2)}] (.8,1.1) ellipse (1.5cm and .3cm);
    \draw[rotate around={59:(-.7,1.2)}] (-.8,1.1) ellipse (1.5cm and .3cm);

@@

    % \tedge{l}{bl}{solid}{}{};
    % \tedge{bl}{b}{solid}{}{};
    % \tedge{r}{br}{solid}{}{};
    % \tedge{br}{b}{solid}{}{};
\end{tikzpicture}

@@
